Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-16 agujournaltemplate.tex
Start writing abstract

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
\index Instead of a unidirectional trend, we see a shift around 2004 to low-upwelling conditions, and a return to the previous community and physical environment in 2014. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, but community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is sea surface temperature (SST), with the MEI v2 index (lag of 4 months) and nitrate concentration following behind. 
\end{keypoints}

@@

\begin{abstract}
The structure of the phytoplankton community was identified using microscopy in the Cariaco basin, in Venezuela, between 1995 and 2017. Microscopy data are limited to larger cell sizes (> 20 µm), where the community showed a large change in the abundances of functional groups and shifts in diversity patterns. 
\end{abstract}

%1
%\section*{Plain Language Summary}
%Enter your Plain Language Summary here or delete this section.
%Here are instructions on writing a Plain Language Summary: 
%https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary

@@

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

>>>>

\begin{keypoints}
\item The CARIACO time series phytoplankton community data reveals a shift and return between two clusters, defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index. 
\index Between the clusters, we observe no significant changes in diversity metrics, except for lower turnover and higher community evenness under low-upwelling conditions.
\item Gradient forest analysis identified the AMO climate index as the strongest predictor of community changes, highlighting the influence of large-scale climatic oscillations on the tropical coastal ecosystem.  
\end{keypoints}

@@

\begin{abstract}
Phytoplankton play a crucial role in Earth's biogeochemical cycles. Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem. 
This study aims to investigate the drivers and effects of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017. Monthly cruises sampled microscopy cell counts, as well as biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis.

- Focus on microscopy data
- Look at general trends and time series of diversity
- Find structure in community data, relate that to shifts in bottom-up drivers
Reveals two clusters, characterized by low and high upwelling. No significant change in diversity, but changes in turnover and evenness.
- Rank predictors of community change using gradient forest analysis
Shows strong impact of upwelling system (i.e. temperature & nitrate), but also climate indices AMO and MEIv.2 are ranked among best predictors of changes in community.

Our results highlight the effect of global climate on a tropcial coastal ecosystem, through decadal scale oscillations.

The structure of the phytoplankton community was identified using microscopy . 
Microscopy data are limited to larger cell sizes (> 20 µm), where the community exhibited significant changes in the the abundances of functional groups and shifts in diversity patterns.

\end{abstract}

%1
\section*{Plain Language Summary}
%Enter your Plain Language Summary here or delete this section.
%Here are instructions on writing a Plain Language Summary: 
%https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary

Microscopic algae drive

@@

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

### 2025-05-16 text-backup.tex
Start writing abstract

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% KEY POINTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences

% Example:
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
% \end{keypoints}

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
\index Instead of a unidirectional trend, we see a shift around 2004 to low-upwelling conditions, and a return to the previous community and physical environment in 2014. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, but community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is sea surface temperature (SST), with the MEI v2 index (lag of 4 months) and nitrate concentration following behind. 
\end{keypoints}
